---
title: "Vignette: How to use ABEILLE"
subtitle: "ABEILLE: a novel method for ABerrant Expression Identification empLoying machine LEarning from sequencing data"
author: "Gwendal Le Bideau, David Pratella, Samira Ait-El-Mkadem - Saadi, Sylvie Bannwarth, Véronique Paquis-Flucklinger, Silvia Bottini"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Vignette: How to use ABEILLE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
In this vignette let us show how to use ABEILLE, and show how the model can be used in a customizable way. The first thing to do is to load the package.
```{r setup}
library(ABEILLE)
```

In the package there is a toy dataset, it can be load with *ExampleAbeilleDataSet*.

```{r}
toy_dataset <- ExampleAbeilleDataSet
print(paste("The dataframe contains",
            dim(toy_dataset)[1],"transcripts and",
            dim(toy_dataset)[2],"samples"))
head(toy_dataset[,1:6])
```
As you can see, ABEILLE is build to work with dataset load as dataframe where transcripts/genes are in rows and samples in columns. The next step is to use the function *DataIntegrity* to check if there is a problem with the data (like NaN or Inf values for example).

```{r}
DataIntegrity(toy_dataset)
```

Since transcript with 0 counts on the entire row is not useful for ABEILLE you can already remove these rows. The function *RemoveZeroCounts* is here for that purpose.

```{r}
#To show the function RemoveZeroCounts let us add a row with only 0
toy_dataset[14001,] <- 0
print(dim(toy_dataset))
toy_dataset <- RemoveZeroCounts(toy_dataset)
```
Let us make a save of this dataset for the next function
```{r eval=FALSE}
write.csv(toy_dataset, "toy_dataset.csv")
#Don't add arguments
```
The goal is now to recreate the data without the outliers. To do so, the dataset needs pass in an autoencoder. We provided a python code containing a VAE named *abeille.py* findable in the package. It can be launched on R using the *reticulate* package or you can use it directly on python. We highly recommend to use it on a machine with GPU acceleration. Feel free to use your own neural network and try to find an advanced architecture.

So let us load *reticulate*, source python and run the loaded function.
```{r eval=FALSE}
library(reticulate)
#Source my own python environment
#To find your python path you can use os.path.dirname(sys.executable) on python
use_python("C:/ProgramData/Anaconda3")
#If you don't have environment ready, you can load library needed in python
py_install("numpy")
py_install("pandas")
py_install("tensorflow")
#Load the VAE function
source_python("abeille.py")
toy_dataset_recons <- abeille_VAE("toy_dataset.csv")
```
Here is the list of all the arguments usable in the *abeille_VAE* function:

WARNING: if you use it on R arguments using an integer needs to be set as 20L for 20 (for example)

* **file** (only mandatory parameter), the path to the file to pass in the autoencoder.
* **logarithm** (default True), boolean to log the input data.
* **read_count** (default True), boolean to correct value bellow 0 (needed if you use read count data).
* **batch_size** (default 30), number of training examples utilized in one iteration.
* **epochs** (default 1500), number of time the dataset is passed forward and backward through the neural network.
* **kernel** (default "lecun_normal"), type of initialization of the neurons.
* **kl_loss_weight** (default 0.5), weight of the Kullback - Leibler loss.
* **edl1** (default 2048), stand for encoder dense layer 1, number of layer in the first hidden layer.
* **edl2** (default 1024), stand for encoder dense layer 2, number of layer in the second hidden layer.
* **edl3** (default 512), stand for encoder dense layer 3, number of layer in the third hidden layer.
* **edl4** (default 256), stand for encoder dense layer 4, number of layer in the fourth hidden layer.
* **latent_size** (default 128), size of the latent space
* **ddl1** (default 256), stand for decoder dense layer 1, number of layer in the sixth layer.
* **ddl2** (default 512), stand for decoder dense layer 2, number of layer in the seventh layer.
* **ddl3** (default 1024), stand for decoder dense layer 3, number of layer in the eigth layer.
* **ddl4** (default 2048), stand for decoder dense layer 4, number of layer in the ninth layer.

Since there is already a reconstruction generate by the ABEILLE VAE available in the package let us load it
```{r}
toy_dataset_recons <- ExampleAbeilleReconstructed
```

To judge the performance of the reconstruction use the function *StatsPred*
```{r}
StatsPred(toy_dataset, toy_dataset_recons)
```

Then two metric is needed for the linear regression, ABEILLE offers a function to compute the Z-score and the Log2 fold-change. As the neural network you can provide your own metric.
```{r}
zscore <- Zscore(toy_dataset, toy_dataset_recons)
l2fc <- L2FC(toy_dataset, toy_dataset_recons)
```

Finally to obtain the outliers use the function *PickOutliers*
```{r}
outliers <- PickOutliers(zscore, l2fc, toy_dataset, toy_dataset_recons)
print(head(outliers))
```
It is possible to create your own decision tree, to pick outliers. This one was build to find outliers following the OUTRIDER outliers distribution.
```{r}
light_tree <- function(linear_reg_data){
  bool_vector <- ifelse(linear_reg_data["cooksD"] >= 15.0,TRUE,FALSE)
  return(bool_vector)
}
outliers <- PickOutliers(zscore, l2fc, toy_dataset, toy_dataset_recons, decision_tree = light_tree)
print(head(outliers))
```

In the package you can also find a function to estimate the size factor. But to use it, you need a table containing information on the samples
```{r}
anno_table <- data.frame(sampleID = colnames(toy_dataset), Sex = sample(c("Male", "Female"), 128, replace = TRUE), Age = sample(1:89, 128, replace = TRUE), Cohort = rep("GTEx database",128))
sizefactor <- SizeFactors(toy_dataset, anno_table)
print(head(sizefactor))
```
Finally there is two function to add outliers inside a dataset. The first on is describe in the OUTRIDER paper.
```{r}
outrider_outliers <- AddOutriderOutliers(toy_dataset, sizefactor, drawoption = 10^-5)
```

The second one is by adding gaussian noise to the counts
```{r}
gaussian_outliers <- AddGaussianOutliers(toy_dataset, drawoption = 10^-5)
```


